---
title: "Learning R with Baseball"
author: "Jackie D. Bonds"
date: "December 31, 2018"
output: html_document
---

```{r, echo=FALSE}
knitr::opts_chunk$set(error = TRUE)
```

```{r packages, include=FALSE}
library(dplyr)
library(ggplot2)
library(Lahman)
```

### Filtering players with greater than 500 at bats

```{r bdata, message=FALSE}
bdata <- Batting %>%
           filter(AB > 500)%>%
           select(playerID, HR, yearID) %>%
           arrange(desc(HR))
head(bdata)
```

### Using & (and) and | (or) to filter with greater detail

```{r bdata2}
bdata2 <- Batting %>%
            filter(AB > 500 & (HR >= 50 | SO < 25))%>%
            select(playerID, HR, yearID) %>%
            arrange(desc(HR))
head(bdata2)
```

### Summing HR data by player

```{r bdata3}
bdata3 <- Batting %>%
            group_by(playerID) %>%
            summarize(career_HR = sum(HR, na.rm = TRUE)) %>%
            filter(career_HR >= 600) %>%
            arrange(desc(career_HR))
head(bdata3)
```

```{r, echo=FALSE}
ggplot(bdata3, aes(x = reorder(playerID, -career_HR), y = career_HR)) + 
  geom_col(fill = "light blue", color = "black") + 
  geom_text(aes(label = career_HR), vjust = 1.25, fontface = "bold")+
  xlab("PlayerID") + 
  ylab("Home_Runs") + 
  ggtitle("All Time Greatest HR Hitters") + 
  theme_classic()
```

### Averaging season hit total by player

```{r bdata4}
bdata4 <- Batting %>%
            group_by(playerID) %>%
            summarize(avg_season_H = round(mean(H, na.rm = TRUE), 2)) %>%
            arrange(desc(avg_season_H))
head(bdata4)
```

### Identifying minimum and maximum HR by player

```{r bdata5}
bdata5 <- Batting %>%
            group_by(playerID) %>%
            summarize(Max_HR = max(HR), Min_HR = min(HR)) %>%
            arrange(desc(Max_HR))
head(bdata5)
```

### Count number of unique records for each player

```{r bdata6}
bdata6 <- Batting %>%
            group_by(playerID) %>%
            summarize(Unique_records = n())
head(bdata6)
```

### Strikeouts by player 

```{r bdata7}
bdata7 <- Batting %>%
            filter(AB >= 400) %>%
            group_by(playerID) %>%
            summarize(Min_SO = min(SO, na.rm = TRUE)) %>%
            filter(Min_SO < 20) %>%
            arrange(Min_SO)
head(bdata7)
```

### Batting AVG by player (best year)

```{r bdata8}
bdata8 <- Batting %>%
            filter(AB >= 400) %>%
            group_by(playerID) %>%
            mutate(Batting_AVG = round(H/AB, 3)) %>%
            select(playerID, Batting_AVG, yearID) %>%
            arrange(desc(Batting_AVG))
head(bdata8)
```

### Batting AVG by player (career)

```{r bdata9}
bdata9 <- Batting %>%
            group_by(playerID) %>%
            summarize(Career_H = sum(H, na.rm = TRUE), Career_AB = sum(AB, na.rm = TRUE)) %>%
            filter(Career_AB >= 1000) %>%
            mutate(Career_AVG = round(Career_H/Career_AB, 3)) %>%
            select(playerID, Career_AVG) %>%
            arrange(desc(Career_AVG))
head(bdata9)
```
  
### Joining the biological information to the batting data

```{r Join}
Join <- inner_join (Batting, Master, by = c("playerID")) %>%
          filter(playerID == "ruthba01" | playerID == "aaronha01") %>%
          arrange(playerID)
head(Join)
```

### Alternative way to join the data

```{r Join2}
data <- Batting %>%
          filter(playerID == "ruthba01" | playerID == "aaronha01")

Join2 <- inner_join(data, Master, by = c("playerID")) %>%
          arrange(playerID)
head(Join2)
```

### Create a new database with player names, team, year, and HR then sort by most HR

```{r Join3}
bdat <- Batting %>%
          select(playerID, teamID, yearID, HR)

Join3 <- inner_join(bdat, Master, by = c("playerID")) %>%
          select(nameFirst, nameLast, teamID, yearID, HR) %>%
          arrange(desc(HR))
head(Join3)
```
           
### Joining batting data with team data (Babe Ruth)

```{r Join4}
bdat <- Batting %>%
          filter(playerID == "ruthba01") %>%
          select(playerID, teamID, yearID, HR)

Join4 <- inner_join(bdat, Teams, by = c("teamID", "yearID")) %>%
          select(playerID, name, yearID, HR = HR.x)
Join4
```

### Joining batting data with Master and team data (Babe Ruth)

```{r Join4.1}
bdat <- Batting %>%
          filter(playerID == "ruthba01") %>%
          select(playerID, teamID, yearID, HR)

Join4 <- inner_join(bdat, Master, by = c("playerID")) %>%
            select(nameFirst, nameLast, teamID, yearID, HR)

Join4.1 <- inner_join(Join4, Teams, by = c("teamID", "yearID")) %>%
              select(nameFirst, nameLast, name, yearID, HR = HR.x)
Join4.1
```

### Summing HR by player while joining Batting with Master

```{r Join5}
bdat <- Batting %>%
          group_by(playerID) %>%
          summarize(career_HR = sum(HR, na.rm = TRUE))

Join5 <- inner_join(bdat, Master, by = c("playerID")) %>%
            select(nameFirst, nameLast, career_HR) %>%
            arrange(desc(career_HR))
head(Join5)
```

#### Alternative method for Join5

```{r Join5.1}
Join5.1 <- inner_join(Batting, Master, by = c("playerID")) %>%
              group_by(playerID) %>%
              summarize(First_name = nameFirst[1], Last_name = nameLast[1], career_HR = sum(HR, na.rm = TRUE)) %>%
              select(First_name, Last_name, career_HR) %>%
              arrange(desc(career_HR))
head(Join5.1)
```

## Project1 <- 
#### Generate a list of all the players who played for a team whose park was Petco Park.  
#### Each such player should be listed with his first and last name.
#### No player should be listed more than once.
#### Limit the list to players in the Batting table.

```{r tdat}
tdat <- Teams %>%
           filter(park == "Petco Park") %>%
           select(teamID, yearID)
```

##### We realize the dates for Petco are 2004 thru 2016 so we can filter directly from the Batting data

```{r Project1}
bdat <- Batting %>%
          filter(teamID == "SDN", yearID >= 2004, yearID <= 2016) %>%
          select(playerID)

Project1 <- inner_join(bdat, Master, by = c("playerID")) %>%
            group_by(playerID) %>%
            summarize(First_name = nameFirst[1], Last_name = nameLast[1]) %>%
            select(First_name, Last_name)
summary(Project1)            
```

## Project2 <- 
#### List all players named "Bob" who averaged more than $1 Million per year in salary for their career.
#### Include last names.

```{r Project2}
bdat <- inner_join(Batting, Master, by = c("playerID")) %>%
              group_by(playerID) %>%
              summarize(First_name = nameFirst[1], Last_name = nameLast[1]) %>%
              select(playerID, First_name, Last_name)

salary <- Salaries %>%
            group_by(playerID) %>%
            summarize(AVG_Salary = mean(salary)) %>%
            filter(AVG_Salary > 1000000) %>%
            select(playerID, AVG_Salary)

Project2 <- inner_join(salary, bdat, by = c("playerID")) %>%
              select(First_name, Last_name, AVG_Salary) %>%
              filter(First_name == "Bob") %>%
              arrange(desc(AVG_Salary))
Project2
```

```{r, echo = FALSE}
ggplot(Project2, aes(reorder(Last_name, -AVG_Salary), AVG_Salary/1e6)) +
  geom_col(fill = "light blue", color = "black") +
  geom_text(aes(label = paste("$",round(AVG_Salary/1000000,2), "M")), vjust = 1.25, fontface = "bold") +
  xlab("Name") + 
  ylab("Salary (Millions)") +
  ggtitle("Bob's Averaging Over $1M Per Year") + 
  theme_classic()  
```  

##### *Much of the learnings were acquired through Udemy's "Baseball Database Queries with SQL and dplyr" taught by Charles Redmond.*  
[link](https://www.udemy.com/baseball1/)
  

